Type-Safe Modular Parsing

### 2016-10-19 paper/sections/Library/DifferentSyntax.tex
Text.

val pET : (=> Pair) => Pair = {
    val pE = pVar2 ||| pApp2 ||| pLam
    val pT = pInt
    p => (pE(p), pT(p))
}
\end{lstlisting}
The last value \lstinline{p} integrates the four single parsers defined before, so we use \lstinline{fix(p)} to
obtain a pair two parsers we want: one for expressions, the other for types. The parser \lstinline{fix(p)._1} parsers expressions
like \lstinline{"\x:Int.x x"} successfully. Yet meanwhile, some drawbacks of this approach come into sight. With different syntax, it
is necessary to extend the fixpoint to contain more parsers, in which case usually some structures like tuples are used,
making code tedious and less elegant. On the other hand, the modularity is quite restricted; only parsing and datatypes are extensible.
Suppose now we want to realize some behaviors on those data structures, for example pretty-printing, then whenever we introduce a new behavior,
it needs implementations in every class body, requiring existing code to be modified. At this point, Object Algebras are adopted for data structures
instead.

Object Algebras are a nice solution to the famous \textit{Expression Problem}, achieving two dimensions of extensibility (data variants and operations)
in a type-safe way. The intuition of using Object Algebras is to separate data structures from behaviors, so as to modularize the design. We can simply add new variants for the new cases in the grammar, or realize different operations on structures, without modifying existing code, and separate compilation is supported. Coincidentally, Object Algebras use generics, where the type parameters perfectly support extensible different syntax, and hence simplify the code comparatively. Below is the code using Object Algebras, where we start from \lstinline{varE} and \lstinline{appE} only, later we will extend them with \lstinline{intT} and \lstinline{lamE}:
\begin{lstlisting}
trait OldAlg[E] {
    def varE(x : String) : E
    def appE(e1 : E, e2 : E) : E
}

trait OldParser[E] {
    val pE : OldAlg[E] => (=> PackratParser[E]) => PackratParser[E] = alg => p =>
        ident ^^ { x => alg.varE(x) } |||
        p ~ p ^^ { case e1 ~ e2 => alg.appE(e1, e2) }
}
\end{lstlisting}
Here \lstinline{OldAlg} is called an \textit{Object Algebra interface}, which has two constructs for the AST. Its parser is
supposed to produce a structure from an input, whereas with Object Algebras structures are implicit: an \textit{algebra}
is immediately applied to the structure and returns the result. It means such a structure that the parser generates should have
type \lstinline{"OldAlg[E] =>} \lstinline{E"} for some \lstinline{E}. In the above code, \lstinline{alg} which has type
\lstinline{OldAlg[E]} is an algebra, and appears as a parameter of \lstinline{pE}, then in the body of \lstinline{pE} the algebra
is invoked correspondingly right after parsing. Since it should be a generic algebra, the parser is enclosed by the generic trait \lstinline{OldParser}.

>>>>

val pET : (=> Pair) => Pair = {
    val pE = pVar2 ||| pApp2 ||| pLam
    val pT = pInt
    p => (pE(p), pT(p))
}
\end{lstlisting}
Based on the old \lstinline{pVar} and \lstinline{pApp}, we implement \lstinline{pVar2} and \lstinline{pApp2} which take the pair as their fix-points instead.
In that case the parsers are consistent, and hence can be composed using the combinator \lstinline{|||}.
The last value \lstinline{pET} integrates the four single parsers defined before, where parsers for expressions and parsers for types are composed respectively. Furthermore, we use \lstinline{fix(p)} to obtain the pair of two parsers we want: one for expressions, the other for types. The parser \lstinline{fix(p)._1} parsers expressions like \lstinline{"\x:Int.x x"} successfully. Yet meanwhile, some drawbacks of this approach come into sight. With different syntax, it
is necessary to extend the fix-point to contain more parsers, in which case usually some structures like tuples are used,
making code tedious and less elegant. Moreover, we need to update such a fix-point in the arguments of all previously defined parsers, so as to make parsers consistent for the combinator (things get even worse without that combinator).
On the other hand, the modularity is quite restricted; only parsing and datatypes are extensible.
Suppose now we want to realize some behaviors on those data structures, for example pretty-printing, then whenever we introduce a new behavior,
it needs implementations in every class body, requiring existing code to be modified. At this point, Object Algebras are adopted for data structures
instead.

One intuition of using Object Algebras is to separate data structures from behaviors, so as to modularize the design, and both parts are even more modular and reusable. We can simply add new variants corresponding to the new cases in the grammar, or realize different operations on structures, without modifying existing code, meanwhile separate compilation is supported. Another one is that Object Algebras benefit us from using generics, where we can abstract different syntax in the type parameters, and define less classes, simplifying code comparatively. Below we present the code using Object Algebras, where we start from \lstinline{varE} and \lstinline{appE} only, later we will extend them with \lstinline{intT} and \lstinline{lamE}:
\begin{lstlisting}
trait OldAlg[E] {
    def varE(x : String) : E
    def appE(e1 : E, e2 : E) : E
}

trait OldParser[E] {
    val pE : OldAlg[E] => (=> Parser[E]) => Parser[E] = alg => p =>
        ident ^^ { x => alg.varE(x) } |||
        p ~ p ^^ { case e1 ~ e2 => alg.appE(e1, e2) }
}
\end{lstlisting}
Here \lstinline{OldAlg} is an Object Algebra interface. An Object Algebra interface is more like a small language, where several cases (constructors) are put together. Since with Object Algebras structures are implicit, a parser of \lstinline{OldAlg} should generate ``structures'' with
type \lstinline{"OldAlg[E] =>} \lstinline{E"} for any \lstinline{E}. In \lstinline{OldParser}, \lstinline{alg} of type
\lstinline{OldAlg[E]} can be any algebra, which appears as a parameter of \lstinline{pE}, then in the body of \lstinline{pE} the algebra
is invoked correspondingly right after parsing. It means we do not get any explicit objects generated from parsers, but they are immediately applied with algebras and return the results. To make \lstinline{alg} as a generic algebra, the enclosing trait of the parser, namely \lstinline{OldParser}, is generic.
